---
title: "Analysis"
author: "Aazar and Nadia"
date: "2024-05-01"
output: pdf_document
---



```{r}
library('dplyr')
data = read.csv("diabetes_prediction_dataset.csv")
data <- data %>%
  mutate(weight_status = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Healthy Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30  ~ "Obesity"
  ))
head(data)
```


# We use a Beta-Binomial Model

+ The choice of alpha and beta is 3 because our prior belief is that theta is 0.5 for all the groups. A randomly chosen person, regardless of any group has a 50 percent chance of having diabetes. 
```{r}
categories = c('Overweight','Obesity', 'Healthy Weight', 'Underweight')

# Create an empty plot
plot(NULL, NULL, xlim=c(0, 0.25), ylim=c(0, 500), xlab="Theta", ylab="Posterior of theta given x", main="Posterior Distributions by Weight Status", type="n")

for (category in categories){
  diabetesunder <- data %>%
    filter(weight_status == category)
  
  data_summary <- table(diabetesunder$weight_status, diabetesunder$diabetes)
  n = sum(data_summary[category, ]) # Total patients underweight
  x = data_summary[category, "1"] 
  
  set.seed(202289)
  
  alpha_prior = 3; beta_prior = 3; 
  theta =seq(0,1,0.00001)
  
  prior = dbeta(theta, alpha_prior, beta_prior)
  
  # likelihood
  likelihood = dbinom(x, n, theta)
  
  # posterior
  alpha_post = alpha_prior + x
  beta_post = beta_prior + n - x
  
  # Plot posterior distribution for each category
  lines(theta, dbeta(theta, shape1=alpha_post, shape2=beta_post), col=rainbow(length(categories))[which(categories == category)], lty=1)
  
  cat( category,": ", qbeta(c(0.025, 0.975), alpha_post, beta_post) , "\n")
  
}

# Add legend
legend("topright", legend=categories, col=rainbow(length(categories)), lty=1, cex=0.8)
```


** Now that we have generated the posterior distributions, it is time to look at the data and see how close the evidence is to the posterior distributions. 
```{r}
data %>%
  group_by(weight_status) %>%
  summarize(
    sum_diabetes = sum(diabetes, na.rm = TRUE),
    percent_diabetic = round((sum_diabetes / n()) * 100,2)
  ) %>%
  arrange(desc(sum_diabetes))
```


+ Since we have a good amount of data, the data plays an overwhelmingly bigger role in determining the posterior distribution.



