---
title: "Analysis"
author: "Aazar and Nadia"
date: "2024-05-01"
output: slidy_presentation
---


```{r, echo= FALSE, message=FALSE}
# Loading in data
library(tidyverse)
library(dplyr)
library(kableExtra)

diabetes <- read_csv("~/Downloads/diabetes_prediction_dataset.csv")

diabetes <- diabetes %>%
  mutate(weight_status = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Healthy Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30  ~ "Obesity"
  ))

# Because there is a column with the new bmi categories, I'm unsure if we still want to keep the bmi values
# diabetes <- diabetes %>% 
#   select(-bmi)

diabetes$hypertension <- factor(diabetes$hypertension, levels = c(0, 1))
diabetes$heart_disease <- factor(diabetes$heart_disease, levels = c(0, 1))
diabetes$smoking_history <- factor(diabetes$smoking_history, levels = c("current", "ever", "former", "never", "No Info", "not current"))
diabetes$diabetes <- factor(diabetes$diabetes, levels = c(0, 1))

table(diabetes$weight_status)
```



# What is in this dataset ?
```{r}
#Neatly presents table
head(diabetes,10) %>%
  kbl() %>%
  kable_paper() %>%
  scroll_box(height = "300px")
```

# Beta Binomial Model (lecture 7)
The analysis revealed a 95% credible interval for the posterior distribution of theta, the proportion of patients with diabetes among those with hypertension, ranging from approximately 0.269 to 0.289. Additionally, the sample proportion of patients with both conditions was estimated to be approximately 0.279. Furthermore, the posterior mean estimate of the proportion of patients with diabetes among those with hypertension was found to be approximately 0.279.
```{r}
library(ggplot2)
# Counting the number of patients with hypertension and diabetes
data_summary <- table(diabetes$hypertension, diabetes$diabetes)
n = sum(data_summary["1", ])  # Total patients with hypertension
x = data_summary["1", "1"]    # Patients with hypertension and diabetes

theta =seq(0,1,0.0001)
alpha_prior = 3
beta_prior = 1
prior = dbeta(theta, alpha_prior, beta_prior)
likelihood = dbinom(x, n, theta)
# posterior
alpha_post = alpha_prior + x
beta_post = beta_prior + n 
#posterior = dbeta(theta, alpha_post, beta_post)
# posterior credible interval
qbeta(c(0.025, 0.975), alpha_post, beta_post)


# Plot posterior distribution for each category
plot(NULL, NULL, xlim=c(0, 0.25), ylim=c(0, 100), xlab="Theta", ylab="Posterior of theta given x", main="Posterior Distribution", type="n")

lines(theta, dbeta(theta, shape1=alpha_post, shape2=beta_post))


paste("The sample proportion is: ", x/n)
```

# Diabetic Gluclose Levels
We’re reasonably confident that the average glucose level lies in the range of 170 to 190 mg/dL. Our prior probability that the average glucose level is between 170 mg/dL and 190 mg/dL is about 68%. Our prior probability that the average glucose level is between 160 mg/dL and 200 mg/dL is about 95%.
```{r}
library(rjags)
library(coda)

set.seed(3153)
tau0 = 10 
mu0 = 180

diabetic <- diabetes %>% 
  filter(diabetes == 1)
sigma
#data 
n= nrow(diabetic) 
sigma= sd(diabetic$blood_glucose_level) 
xbar= mean(diabetic$blood_glucose_level)

x = diabetic$blood_glucose_level
#posterior SD
tau_n = 1 / sqrt(n / sigma ^ 2 + 1 / tau0 ^ 2)
#  0.6347694
tau_n
tau_n
#posterior mean
mu_n = mu0 * (1 / tau0 ^ 2) / (1 / tau_n ^ 2) + xbar * (n / sigma ^ 2) / (1 / tau_n ^ 2)
theta = seq(mu0 - 3 * tau0, mu0 + 3 * tau0, 0.01) # the grid is just for plotting
# 194.0379
mu_n
# prior
prior = dnorm(theta, mu0, tau0)
# likelihood
likelihood = dnorm(xbar, theta, sigma / sqrt(n))
# posterior
posterior = dnorm(theta, mu_n, tau_n)
# plot
ymax = max(c(prior, posterior))
scaled_likelihood = likelihood * ymax / max(likelihood)

plot(theta, prior, type='l', col='red', xlim= range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, scaled_likelihood, type='l', col='blue', xlim=range(theta), ylim=c(0, ymax), ylab='',  yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, posterior, type='l', col='seagreen', xlim=range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
legend("topleft", c("prior", "scaled likelihood", "posterior"), lty=1, col=c("red", " blue", "green"),lwd=3,  bty = "n")
```
# There is a 95% posterior probability that the true mean glucose level for the diabetic population falls between 192.794 mg/dL and 195.282 mg/dL.
```{r}
qnorm(c(0.025,0.975), 194.0379,  0.6347694) # Postereior Mean
```

# 
```{r}
library(rjags)
Nrep = 10000
Nchains = 3

# data
# data has already been loaded in previous code

# model
model_string <- "model{

  # Likelihood
  for (i in 1:n){
    x[i] ~ dnorm(theta, 1 / sigma ^ 2)
  }
  sigma <-   58.64111
  
  #OR simply dnorm(theta, n / sigma ^ 2)
  

  # Prior
  theta ~ dnorm(mu0, 1 / tau0 ^ 2)
  mu0 <- 180
  tau0 <- 10

}"

# Compile the model
dataList = list(x=x, n=n)

model <- jags.model(textConnection(model_string), 
                    data=dataList,
                    n.chains=Nchains)


## --------------------------------
update(model, 1000, progress.bar="none")

posterior_sample <- coda.samples(model, 
                                 variable.names=c("theta"),
                                 n.iter=Nrep,
                                 progress.bar="none")
# Summarize and check diagnostics
summary(posterior_sample)


## --------------------------------
plot(posterior_sample)
```

# 
```{r}
## ---- fig.height=3---------------
theta_sim = rnorm(100000, mu_n, tau_n)
x_sim = rnorm(100000, theta_sim, sigma)
hist(x_sim, freq = FALSE, breaks = 30,
 xlab = "Diabetic Blood Glucose Level",
 main = "Posterior predictive distribution")
```

#
This interval suggests that we are 95% confident that the glucose levels of a diabetic will fall between 79.2 and 308.2 mg/dL. This range captures the variability in glucose levels that can be expected among individuals with diabetes, from fairly low levels that are near normal to significantly elevated levels that indicate poor glucose control.
```{r}
## --------------------------------
quantile(x_sim, c(0.025, 0.975))
#    2.5%    97.5% 
# 79.1999 308.1716 

## --------------------------------
(sum(x < quantile(x_sim, 0.025)) + sum(x > quantile(x_sim, 0.975))) / n

```

# Non-Diabetic Gluclose Levels 
```{r}
library(rjags)
library(coda)

set.seed(3153)
tau0 = 10 
mu0 = 100

non_diabetic <- diabetes %>% 
  filter(diabetes == 0)

#data 
n= nrow(non_diabetic) 
sigma= sd(non_diabetic$blood_glucose_level) 
xbar= mean(non_diabetic$blood_glucose_level)
sigma
x <- non_diabetic$blood_glucose_level
#posterior SD
tau_n = 1 / sqrt(n / sigma^2 + 1 / tau0^2)
theta = seq(mu0 - 3 * tau0, mu0 + 3 * tau0, 0.01) # the grid is just for plotting
#  0.1132108
tau_n
#posterior mean

mu_n = mu0 * (1 / tau0 ^ 2) / (1 / tau_n ^ 2) + xbar * (n / sigma ^ 2) / (1 / tau_n ^ 2)

 # 132.8483

# prior
prior = dnorm(theta, mu0, tau0)
# likelihood
likelihood = dnorm(xbar, theta, sigma / sqrt(n))
# posterior
posterior = dnorm(theta, mu_n, tau_n)

# plot
ymax = max(c(prior, posterior))
scaled_likelihood = likelihood * ymax / max(likelihood)

plot(theta, prior, type='l', col='red', xlim= range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, scaled_likelihood, type='l', col='blue', xlim=range(theta), ylim=c(0, ymax), ylab='',  yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, posterior, type='l', col='seagreen', xlim=range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
legend("topleft", c("prior", "scaled likelihood", "posterior"), lty=1, col=c("red", " blue", "green"),lwd=3,  bty = "n")

qnorm(c(0.025,0.975), 132.8483, .1132108)
```

```{r}
library(rjags)
Nrep = 10000
Nchains = 3

# data
# data has already been loaded in previous code

# model
model_string <- "model{

  # Likelihood
  for (i in 1:n){
    x[i] ~ dnorm(theta, 1 / sigma ^ 2)
  }
  sigma <-   34.24728
  
  #OR simply dnorm(theta, n / sigma ^ 2)
  

  # Prior
  theta ~ dnorm(mu0, 1 / tau0 ^ 2)
  mu0 <- 100
  tau0 <- 10

}"

# Compile the model
dataList = list(x=x, n=n)

model <- jags.model(textConnection(model_string), 
                    data=dataList,
                    n.chains=Nchains)


## --------------------------------
update(model, 1000, progress.bar="none")

posterior_sample <- coda.samples(model, 
                                 variable.names=c("theta"),
                                 n.iter=Nrep,
                                 progress.bar="none")
# Summarize and check diagnostics
summary(posterior_sample)


## --------------------------------
plot(posterior_sample)
```
#
```{r}
## ---- fig.height=3---------------
theta_sim = rnorm(100000, mu_n, tau_n)
x_sim = rnorm(100000, theta_sim, sigma)
hist(x_sim, freq = FALSE, breaks = 30,
 xlab = "Non- Diabetic Blood Glucose Level",
 main = "Posterior predictive distribution")
```
# This interval suggests that we are 95% confident that the glucose levels of a diabetic will fall between 65.90 and 200.03 mg/dL.
```{r}
## --------------------------------
quantile(x_sim, c(0.025, 0.975))


## --------------------------------
(sum(x < quantile(x_sim, 0.025)) + sum(x > quantile(x_sim, 0.975))) / n

```

# We use a Beta-Binomial Model

```{r}
library('dplyr')
data = read.csv("diabetes_prediction_dataset.csv")
data <- data %>%
  mutate(weight_status = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Healthy Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30  ~ "Obesity"
  ))
head(data)
```

The choice of alpha and beta is 3 because our prior belief is that theta is 0.5 for all the groups. A randomly chosen person, regardless of any group has a 50 percent chance of having diabetes. 
```{r}
categories = c('Overweight','Obesity', 'Healthy Weight', 'Underweight')

# Create an empty plot
plot(NULL, NULL, xlim=c(0, 0.25), ylim=c(0, 500), xlab="Theta", ylab="Posterior of theta given x", main="Posterior Distributions by Weight Status", type="n")

for (category in categories){
  diabetesunder <- data %>%
    filter(weight_status == category)
  
  data_summary <- table(diabetesunder$weight_status, diabetesunder$diabetes)
  n = sum(data_summary[category, ]) # Total patients underweight
  x = data_summary[category, "1"] 
  
  set.seed(202289)
  
  alpha_prior = 3; beta_prior = 3; 
  theta =seq(0,1,0.00001)
  
  prior = dbeta(theta, alpha_prior, beta_prior)
  
  # likelihood
  likelihood = dbinom(x, n, theta)
  
  # posterior
  alpha_post = alpha_prior + x
  beta_post = beta_prior + n - x
  
  # Plot posterior distribution for each category
  lines(theta, dbeta(theta, shape1=alpha_post, shape2=beta_post), col=rainbow(length(categories))[which(categories == category)], lty=1)
  
  cat( category,": ", qbeta(c(0.025, 0.975), alpha_post, beta_post) , "\n")
  
}

# Add legend
legend("topright", legend=categories, col=rainbow(length(categories)), lty=1, cex=0.8)
```


** Now that we have generated the posterior distributions, it is time to look at the data and see how close the evidence is to the posterior distributions. 
```{r}
data %>%
  group_by(weight_status) %>%
  summarize(
    sum_diabetes = sum(diabetes, na.rm = TRUE),
    percent_diabetic = round((sum_diabetes / n()) * 100,2)
  ) %>%
  arrange(desc(sum_diabetes))
```


+ Since we have a good amount of data, the data plays an overwhelmingly bigger role in determining the posterior distribution.





