---
title: "Bayesian Insights into Diabetes"
author: "Aazar and Nadia"
date: "2024-04-28"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo= FALSE}
# Loading in data
library(tidyverse)
library(dplyr)
library(kableExtra)

diabetes <- read_csv("diabetes_prediction_dataset.csv")

diabetes <- diabetes %>%
  mutate(weight_status = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Healthy Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30  ~ "Obesity"
  ))

# Because there is a column with the new bmi categories, I'm unsure if we still want to keep the bmi values
# diabetes <- diabetes %>% 
#   select(-bmi)

diabetes <- read_csv("~/Downloads/diabetes_prediction_dataset.csv")
diabetes$hypertension <- factor(diabetes$hypertension, levels = c(0, 1))
diabetes$heart_disease <- factor(diabetes$heart_disease, levels = c(0, 1))
diabetes$smoking_history <- factor(diabetes$smoking_history, levels = c("current", "ever", "former", "never", "No Info", "not current"))
diabetes$diabetes <- factor(diabetes$diabetes, levels = c(0, 1))

```
# What is in this dataset ?
```{r}
#Neatly presents table
head(diabetes,10) %>%
  kbl() %>%
  kable_paper() %>%
  scroll_box(height = "300px")
```



# Beta Binomial Model (lecture 7)
The analysis revealed a 95% credible interval for the posterior distribution of theta, the proportion of patients with diabetes among those with hypertension, ranging from approximately 0.269 to 0.289. Additionally, the sample proportion of patients with both conditions was estimated to be approximately 0.279. Furthermore, the posterior mean estimate of the proportion of patients with diabetes among those with hypertension was found to be approximately 0.279.
```{r}
library(ggplot2)
# Counting the number of patients with hypertension and diabetes
data_summary <- table(diabetes$hypertension, diabetes$diabetes)
n = sum(data_summary["1", ])  # Total patients with hypertension
x = data_summary["1", "1"]    # Patients with hypertension and diabetes

theta =seq(0,1,0.0001)
alpha_prior = 3
beta_prior = 1
prior = dbeta(theta, alpha_prior, beta_prior)
likelihood = dbinom(x, n, theta)
# posterior
alpha_post = alpha_prior + x
beta_post = beta_prior + n 
#posterior = dbeta(theta, alpha_post, beta_post)
# posterior credible interval
qbeta(c(0.025, 0.975), alpha_post, beta_post)

paste("The sample proportion is: ", x/n)
paste("The posterior mean : ",(x+alpha_prior+beta_prior)/ (n+alpha_prior+beta_prior))
```

Lecture 11
Diabetic Gluclose Levels
```{r}
library(rjags)
library(coda)

set.seed(3153)
tau0 = 10 
mu0 = 180

diabetic <- diabetes %>% 
  filter(diabetes == 1)

#data 
n= nrow(diabetic) 
sigma= sd(diabetic$blood_glucose_level) 
xbar= mean(diabetic$blood_glucose_level)

x <- rnorm(n, xbar, sigma)
#posterior SD
tau_n = 1 / sqrt(n / sigma ^ 2 + 1 / tau0 ^ 2)
#  0.6347694

tau_n
#posterior mean
mu_n = mu0 * (1 / tau0 ^ 2) / (1 / tau_n ^ 2) + xbar * (n / sigma ^ 2) / (1 / tau_n ^ 2)
theta = seq(mu0 - 3 * tau0, mu0 + 3 * tau0, 0.01) # the grid is just for plotting
# 194.0379

# prior
prior = dnorm(theta, mu0, tau0)
# likelihood
likelihood = dnorm(xbar, theta, sigma / sqrt(n))
# posterior
posterior = dnorm(theta, mu_n, tau_n)

# plot
ymax = max(c(prior, posterior))
scaled_likelihood = likelihood * ymax / max(likelihood)

plot(theta, prior, type='l', col='red', xlim= range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, scaled_likelihood, type='l', col='blue', xlim=range(theta), ylim=c(0, ymax), ylab='',  yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, posterior, type='l', col='seagreen', xlim=range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
legend("topleft", c("prior", "scaled likelihood", "posterior"), lty=1, col=c("red", " blue", "green"),lwd=3,  bty = "n")

qnorm(c(0.025,0.975), 194.0379,  0.6347694) # Postereior Mean
```

```{r}
library(rjags)
Nrep = 10000
Nchains = 3

# data
# data has already been loaded in previous code

# model
model_string <- "model{

  # Likelihood
  for (i in 1:n){
    x[i] ~ dnorm(theta, 1 / sigma ^ 2)
  }
  sigma <-   58.64111
  
  #OR simply dnorm(theta, n / sigma ^ 2)
  

  # Prior
  theta ~ dnorm(mu0, 1 / tau0 ^ 2)
  mu0 <- 150
  tau0 <- 10

}"

# Compile the model
dataList = list(x=x, n=n)

model <- jags.model(textConnection(model_string), 
                    data=dataList,
                    n.chains=Nchains)


## --------------------------------
update(model, 1000, progress.bar="none")

posterior_sample <- coda.samples(model, 
                                 variable.names=c("theta"),
                                 n.iter=Nrep,
                                 progress.bar="none")
# Summarize and check diagnostics
summary(posterior_sample)


## --------------------------------
plot(posterior_sample)


## ---- fig.height=3---------------
theta_sim = rnorm(100000, mu_n, tau_n)
x_sim = rnorm(100000, theta_sim, sigma)
hist(x_sim, freq = FALSE, breaks = 30,
 xlab = "diabetic Blood Glucose Level",
 main = "Posterior predictive distribution")


## --------------------------------
quantile(x_sim, c(0.025, 0.975))


## --------------------------------
(sum(x < quantile(x_sim, 0.025)) + sum(x > quantile(x_sim, 0.975))) / n

```

Non- Diabetic Gluclose Levels 

```{r}
library(rjags)
library(coda)

set.seed(3153)
tau0 = 10 
mu0 = 100

non_diabetic <- diabetes %>% 
  filter(diabetes == 0)

#data 
n= nrow(non_diabetic) 
sigma= sd(non_diabetic$blood_glucose_level) 
xbar= mean(non_diabetic$blood_glucose_level)

x <- rnorm(n, xbar, sigma)
#posterior SD
tau_n = 1 / sqrt(n / sigma ^ 2 + 1 / tau0 ^ 2)
#  0.1132108
tau_n
#posterior mean
mu_n = mu0 * (1 / tau0 ^ 2) / (1 / tau_n ^ 2) + xbar * (n / sigma ^ 2) / (1 / tau_n ^ 2)
theta = seq(mu0 - 3 * tau0, mu0 + 3 * tau0, 0.01) # the grid is just for plotting
 # 132.8483

# prior
prior = dnorm(theta, mu0, tau0)
# likelihood
likelihood = dnorm(xbar, theta, sigma / sqrt(n))
# posterior
posterior = dnorm(theta, mu_n, tau_n)

# plot
ymax = max(c(prior, posterior))
scaled_likelihood = likelihood * ymax / max(likelihood)

plot(theta, prior, type='l', col='red', xlim= range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, scaled_likelihood, type='l', col='blue', xlim=range(theta), ylim=c(0, ymax), ylab='',  yaxt='n',lwd=3, frame.plot = FALSE)
par(new=T)
plot(theta, posterior, type='l', col='seagreen', xlim=range(theta), ylim=c(0, ymax), ylab='', yaxt='n',lwd=3, frame.plot = FALSE)
legend("topleft", c("prior", "scaled likelihood", "posterior"), lty=1, col=c("red", " blue", "green"),lwd=3,  bty = "n")

qnorm(c(0.025,0.975), 132.8483, .1132108)
```

```{r}
library(rjags)
Nrep = 10000
Nchains = 3

# data
# data has already been loaded in previous code

# model
model_string <- "model{

  # Likelihood
  for (i in 1:n){
    x[i] ~ dnorm(theta, 1 / sigma ^ 2)
  }
  sigma <-  34.24728
  
  #OR simply dnorm(theta, n / sigma ^ 2)
  

  # Prior
  theta ~ dnorm(mu0, 1 / tau0 ^ 2)
  mu0 <- 100  
  tau0 <- 10

}"

# Compile the model
dataList = list(x=x, n=n)

model <- jags.model(textConnection(model_string), 
                    data=dataList,
                    n.chains=Nchains)


## --------------------------------
update(model, 1000, progress.bar="none")

posterior_sample <- coda.samples(model, 
                                 variable.names=c("theta"),
                                 n.iter=Nrep,
                                 progress.bar="none")
# Summarize and check diagnostics
summary(posterior_sample)


## --------------------------------
plot(posterior_sample)


## ---- fig.height=3---------------
theta_sim = rnorm(100000, mu_n, tau_n)
x_sim = rnorm(100000, theta_sim, sigma)
hist(x_sim, freq = FALSE, breaks = 30,
 xlab = "Non-diabetic Blood Glucose Level",
 main = "Posterior predictive distribution")


## --------------------------------
quantile(x_sim, c(0.025, 0.975))


## --------------------------------
(sum(x < quantile(x_sim, 0.025)) + sum(x > quantile(x_sim, 0.975))) / n


```

